Data Specialization for a Staged Lambda Calculus

### 2015-02-05 paper/splitting.tex
tiny edits to section 4

The primary strength of \lang\ is the expressive power afforded by
its ability to syntactically interleave work intended for stage \bbone\
and work intended for stage \bbtwo, as well as it ability to form
mixed-stage abstractions.

In this section, we provide an algorithm to translate the multistage terms and values of \lang\
into equivalent but simpler forms where the stage \bbone\ and stage \bbtwo\ portions 
of the computation are {\em syntactically seperated} from each other.
That is, we desire a outputs of the form $\mathcal{C}(e_1;e_2)$ where  
$e_1$ is a monostage term that encodes all stage \bbone\ computation, 
and $e_2$ is a monostage term that encodes all stage \bbtwo\ computation.
To represent outputs in this way, we introduce three new constructs:
\begin{itemize}
\item {\em Single-output pipelines}, \texttt{pipeS}, for the output of splitting \lang\ terms at world \bbtwo.
\item {\em Multi-output pipelines}, \texttt{pipeM}, for the output of splitting \lang\ terms at world \bbonem. 
\item {\em Masked values}, \texttt{mv}, which will serve a purpose analogous to \lang\ partial values.
\end{itemize}

@@

With these output terms defined, we can state the goal of our splitting algorithm:
to translate terms in \lang\ to a syntactically-separted form, 
such that the semantics is preserved.

Exactly what is meant by ``such that the semantics is preserved'' depends on whether we're translating
terms at \bbonem\ or \bbtwo.  
In the latter case, where an $\coltwo e A$ is translated to a $\pipeS p l r$ with the $\splittwosym$ relation,
this constraint is straightforward: $e$ and $\pipeS p l r$ can both be reduced to residuals
via $\redtwosym$ and $\sepredtwosym$, so we only need to require that those
residuals be equivalent for the translation to be correct.
In the former case, where an $\colone e A$ is translated to a $\pipeM c l r$ with the $\splitonesym$ relation, 
things are more complicated: 
the output of $\redonesym$ is a partial value and the output of $\sepredonesym$ is a masked value,
so we first need a way to relate those forms if we want to talk about correctness of the term translation.
That is the purpose of of the {\em masking} relation, written $\vsplito$.

This section proceeds by first defining masking ($\vsplito$), 
then using it to motivate the definition of \bbonem-translation ($\splitonesym$),
and finally coming back to define \bbtwo-tranlastion ($\splittwosym$).
A summary of these operations and the relationship between them is provided in \ref{fig:splittingSummary}.

@@

What we desire is a simple way to translate this partial value into a form that separates the stage~\bbone\ content (the value \texttt{3}) 
from the stage~\bbtwo\ content (the term $\letin \yhat {\mathtt{7+5}} \yhat$),
allowing us to later define a term translation that respects this value translation.

@@

Since lambdas may represent multi-stage computations, 
masking splits the body of lambdas as a general stage~\bbone\ terms (as described in \ref{sec:split-one}), 
and packages the resulting terms as functions.

>>>>

The primary strength of \lang\ is the expressive power afforded by
its ability to syntactically interleave work intended for stage \bbone\
and work intended for stage \bbtwo, as well as its ability to form
mixed-stage abstractions.

In this section, we provide an algorithm to translate the multistage terms and values of \lang\
into equivalent but simpler forms where the stage \bbone\ and stage \bbtwo\ portions 
of the computation are {\em syntactically separated} from each other.
That is, we desire outputs of the form $\mathcal{C}(e_1;e_2)$ where  
$e_1$ is a monostage term encoding all stage \bbone\ computation, 
and $e_2$ is a monostage term encoding all stage \bbtwo\ computation.
To represent outputs in this way, we introduce three new constructs:
\begin{itemize}
\item {\em Single-output pipelines}, \texttt{pipeS}, for the output of splitting \lang\ terms at world \bbtwo.
\item {\em Multi-output pipelines}, \texttt{pipeM}, for the output of splitting \lang\ terms at world \bbonem. 
\item {\em Masked values}, \texttt{mv}, which will serve a purpose analogous to \lang\ partial values.
\end{itemize}

@@

With these output terms defined, we can state the goal of our splitting algorithm:
to translate terms in \lang\ to a syntactically-separated form, 
such that the semantics is preserved.

Exactly what is meant by ``such that the semantics is preserved'' depends on whether we're translating
terms at \bbonem\ or \bbtwo.  
In the latter case, where an $\coltwo e A$ is translated to a $\pipeS p l r$ with the $\splittwosym$ relation,
this constraint is straightforward: $e$ and $\pipeS p l r$ can both be reduced to residuals
via $\redtwosym$ and $\sepredtwosym$, so we only need to require that those
residuals be equivalent for the translation to be correct.
In the former case, where an $\colone e A$ is translated to a $\pipeM c l r$ with the $\splitonesym$ relation, 
things are more complicated: 
the output of $\redonesym$ is a partial value and the output of $\sepredonesym$ is a masked value,
so we first need a way to relate those forms if we want to talk about correctness of the term translation.
That is the purpose of the {\em masking} relation, written $\vsplito$.

This section proceeds by defining masking ($\vsplito$), 
then using masking to motivate the definition of \bbonem-translation ($\splitonesym$),
and finally coming back to define \bbtwo-translation ($\splittwosym$).
A summary of these operations and the relationship between them is provided in \ref{fig:splittingSummary}.

@@

We want a simple way to translate this partial value into a form that separates the stage~\bbone\ content (the value \texttt{3}) 
from the stage~\bbtwo\ content (the term $\letin \yhat {\mathtt{7+5}} \yhat$),
allowing us to later define a term translation that respects this value translation.

@@

Since lambdas may represent multi-stage computations, 
masking splits the body of lambdas as general stage~\bbone\ terms (as described in \ref{sec:split-one}), 
and packages the resulting terms as functions.
